Generating Random Logic Programs Using Constraint Programming

### 2020-07-12 text/paper2/paper.tex
arXiv compatibility
Finishing touches

\documentclass[runningheads]{llncs}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage[capitalise]{cleveref}
\usepackage{tikz}
\usepackage{mathrsfs}
\usepackage[nounderscore]{syntax}
\usepackage{blkarray}
\usepackage[binary-units]{siunitx}
\usepackage[inline,shortlabels]{enumitem}
\usepackage{capt-of}
\usepackage[caption=false]{subfig}
\usepackage{booktabs}

\captionsetup[subfigure]{width=\linewidth}

\renewcommand\fbox{\fcolorbox{red}{white}}
\newcommand{\hilight}[1]{\setlength{\fboxsep}{1pt}\colorbox{lightgray}{#1}}
\newcommand{\hlitem}{\stepcounter{enumi}\item[\hilight{\theenumi}]}

@@

\makeatletter
\newcommand{\nosemic}{\renewcommand{\@endalgocfline}{\relax}}% Drop semi-colon ;
\newcommand{\dosemic}{\renewcommand{\@endalgocfline}{\algocf@endline}}% Reinstate semi-colon ;
\newcommand{\pushline}{\Indp}% Indent
\newcommand{\popline}{\Indm\dosemic}% Undent
\makeatother

\spnewtheorem{constraint}{Constraint}{\bfseries}{\itshape}

@@

\usetikzlibrary{arrows.meta}
%\relpenalty=10000
%\binoppenalty=10000

@@

\title{Generating Random Logic Programs Using Constraint Programming}
%\author{Paulius Dilkas\inst{1} \and
%Vaishak Belle\inst{1,2}}
%\authorrunning{P. Dilkas and V. Belle}
%\institute{University of Edinburgh, Edinburgh, UK \email{\{p.dilkas@sms.,vaishak@\}ed.ac.uk} \and Alan Turing Institute, London, UK}

@@

  \keywords{Constraint programming \and Probabilistic logic programming \and
    Statistical relational learning.}
\end{abstract}

@@

In the world of CSPs, we also have \emph{(constraint) variables}, each with a
\emph{domain}, whose values are restricted using \emph{constraints}. All
constraint variables in the model are integer or set variables, however, if an
integer refers to a logical construct (e.g., a logical variable or a constant),
we will make no distinction between the two and often use names of logical
constructs to refer to the underlying integers. We say that a constraint
variable is \emph{(fully) determined} if its domain (at the given moment in the
execution) has exactly one value. We will often use $\Box$ as a special domain
value to indicate a `disabled' (i.e., fixed and ignored) part of the model. We
write $\arrayd{a}{b}{c}$ to mean that $\variable{a}$ is an array of variables of
length $b$ such that each element of $\variable{a}$ has domain $c$. Similarly,
we write $\arrayt{a}{b}{c}$ to denote an array $\variable{a}$ of length $b$ such
that each element of $\variable{a}$ has type $\variable{c}$. Finally, we assume
that all arrays start with index zero.

\paragraph{Parameters of the model.} We begin defining the parameters of our
model by initialising sets and lists of the primitives used in constructing
logic programs: a list of predicates $\predicates{}$, a list of their
corresponding arities $\arities{}$ (so $|\arities{}|$ = $|\predicates{}|$), a
set of variables $\variables{}$, and a set of constants $\constants{}$. Either
$\variables{}$ or $\constants{}$ can be empty, but we assume that
$|\constants{}| + |\variables{}| > 0$. Similarly, the model supports zero-arity
predicates but requires at least one predicate to have non-zero arity. For
notational convenience, we also set $\maxArity{} = \max \arities{}$. Next, we
need a measure of how complex a body of a clause can be. As we represent each
body by a tree (see \cref{sec:bodies}), we set $\maxNumNodes{} \ge 1$ to
be the maximum number of nodes in the tree representation of any clause. We also
set $\maxNumClauses{}$ to be the maximum number of clauses in a program. We must
have that $\maxNumClauses{} \ge |\predicates{}|$ because we require each
predicate to have at least one clause that defines it. The model supports
enforcing predicate independence (see \cref{sec:independence}), so a set of
independent pairs of predicates is another parameter. Since this model can
generate probabilistic as well as non-probabilistic programs, each clause is
paired with a probability which is randomly selected from a given list---our
last parameter. For generating non-probabilistic programs, one can set this list
to $[1]$. Finally, we define $\tokens{} = \{ \neg, \land, \lor, \top \}$ as the
set of tokens that (together with atoms) form a clause. All decision variables
of the model can now be divided into $2 \times \maxNumClauses{}$ separate
groups, treating the body and the head of each clause separately. We say that
the variables are contained in two arrays:
$\arrayt{bodies}{\maxNumClauses{}}{Body}$ and
$\arrayt{heads}{\maxNumClauses{}}{Head}$.

@@

We define the \emph{head} of a clause as a $\variable{predicate} \in \predicates{}
\cup \{ \Box \}$ and $\arrayd{arguments}{\maxArity{}}{\constants{} \cup
  \variables{}} \cup \{ \Box \}$. Here, we use $\Box$ to denote either a
disabled clause that we choose not to use or disabled arguments if the arity of
the $\variable{predicate}$ is less than $\maxArity{}$. The reason why we need a
separate value for the latter (i.e., why it is not enough to fix disabled
arguments to a single already-existing value) will become clear in
\cref{sec:variable_symmetry}. This $\variable{predicate}$ variable has a
corresponding arity that depends on the $\variable{predicate}$. We can
define $\variable{arity} \in [0, \maxArity{}]$ as the arity of the
$\variable{predicate}$ if $\variable{predicate} \in \predicates{}$ and zero
otherwise using the table constraint \cite{DBLP:conf/cpaior/MairyDL15}. This
constraint uses a set of pairs of the form $(p, a)$, where $p$ ranges over all
possible values of the $\variable{predicate}$, and $a$ is either the arity of
predicate $p$ or zero. Having defined arity, we can now fix the superfluous
arguments.
\begin{constraint} \label{constr:arity}
  For $i = 0, \dots, \maxArity{} - 1$, $\variable{arguments}[i] = \Box \iff i
  \ge \variable{arity}$.
\end{constraint}
We can also add a constraint that each predicate $\mathsf{P} \in \predicates{}$
should have at least one clause with $\mathsf{P}$ at its head.
\begin{constraint}
  Let $P = \{ h.\variable{predicate} \mid h \in \variable{heads} \}$ be a
  multiset. Then
  \[
    \variable{nValues}(P) = \begin{cases}
      |\predicates{}| & \text{if } \variable{count}(\Box, P) = 0 \\
      |\predicates{}| + 1 & \text{otherwise,}
    \end{cases}
  \]
  where $\variable{nValues}(P)$ counts the number of unique values in $P$, and
  $\variable{count}(\Box, P)$ counts how many times $\Box$ appears in $P$.
\end{constraint}
Finally, clauses are supposed to constitute a set but with one important
exception: there may be more than one disabled clause, i.e., a clause with head
$\variable{predicate} = \Box$. Assuming a lexicographic order over entire
clauses such that $\Box > \mathsf{P}$ for all $\mathsf{P} \in \predicates{}$ and
the head predicate is the `first digit' of this representation, we get the
following constraint.
\begin{constraint}
  For $i = 1, \dots, \maxNumClauses{} - 1$, if $\variable{heads}[i].\variable{predicate}
  \ne \Box$, then
  \[
    (\variable{heads}[i-1], \variable{bodies}[i-1]) < (\variable{heads}[i],
    \variable{bodies}[i]).
  \]
\end{constraint}

@@

To demonstrate the correctness of the model, in this section we are going to
derive combinatorial expressions for counting the number of programs with up to
$\maxNumClauses{}$ clauses and up to $\maxNumNodes{}$ nodes per clause, and
arbitrary $\predicates{}$, $\arities{}$, $\variables{}$, and $\constants{}$.
Being able to establish two ways to generate the same sequence of numbers (i.e.,
numbers of programs with certain properties and parameters) allows us to gain
confidence that the constraint model accurately matches our intentions. For this
section, we introduce the term \emph{total arity} of a body of a clause to refer
to the sum total of arities of all predicates in the body.

@@

Let us order the elements of $\predicates{}$, and let $a_i$ be the arity of the
$i$th predicate. The number of programs is then:
\[
  \sum_{\substack{ \sum_{i=1}^{|\predicates{}|} h_i = n,\\
      |\predicates{}| \le n \le \maxNumClauses{},\\
      h_i \ge 1 \text{ for all } i}} \prod_{i=1}^{|\predicates{}|}
  \binom{\sum_{a=0}^{\maxArity{} \times \maxNumNodes{}} C(a) P(a+a_i)}{h_i},
\]
Here, we sum over all possible ways to distribute $|\predicates{}| \le n \le
\maxNumClauses{}$ clauses among $|\predicates{}|$ predicates so that each
predicate gets at least one clause. For each predicate, we can then count the
number of ways to select its clauses out of all possible clauses. The number of
possible clauses can be computed by considering each possible arity $a$, and
multiplying the number of `unfinished' clauses $C(a)$ by the number of ways to
select the required $a+a_i$ terms in the body and the head of the clause.

@@

In this section, we present custom constraints for stratification and
independence. Stratification is a condition necessary for probabilistic logic
programs \cite{DBLP:conf/padl/MantadelisR17} and often enforced on logic
programs \cite{DBLP:journals/tcs/Bidoit91} that helps to ensure a unique answer
to every query, while independence is a fundamental concept in probability
theory that imposes structure on the probability distribution and simplifies
inference. In the context of probabilistic logic programming languages, these
two seemingly disparate concepts can be defined using the same building block,
i.e., a predicate dependency graph. Constraints are usually presented with two
algorithms: one for \emph{entailment} checking, i.e., checking if the (partially
solved) constraint model satisfies the constraint or not, and one for
\emph{propagation}, i.e., removing values from domains that are incompatible
with the current state of the model. We present both algorithms for independence
while settling for a short description of entailment for stratification.

Let $\mathscr{P}$ be a probabilistic logic program with its set of predicates
$\predicates{}$. Its \emph{(predicate) dependency graph} is a directed graph
$G_{\mathscr{P}}$ with elements of $\predicates{}$ as nodes and an edge between
$\mathsf{P}, \mathsf{Q} \in \predicates{}$ if there is a clause in $\mathscr{P}$
with $\mathsf{Q}$ as the head and $\mathsf{P}$ mentioned in the body. We say
that the edge is \emph{negative} if there exists a clause with $\mathsf{Q}$ as
the head and at least one instance of $\mathsf{P}$ at the body such that the
path from the root to the $\mathsf{P}$ node in the tree representation of the
clause passes through at least one negation node; otherwise, it is
\emph{positive}. We say that $\mathscr{P}$ (or $G_{\mathscr{P}}$) has a
\emph{negative cycle} if $G_{\mathscr{P}}$ has a cycle with at least one
negative edge. A program $\mathscr{P}$ is \emph{stratified} if $G_{\mathscr{P}}$
has no negative cycles.\footnote{This definition is an extension of a well-known
  result for logic programs \cite{DBLP:journals/jlp/BalbinPRM91} to
  probabilistic logic programs with arbitrary complex clause bodies.} Thus a
simple entailment algorithm for stratification can be constructed by selecting
all clauses, all predicates of which are fully determined, and looking for
negative cycles in the dependency graph constructed based on those clauses using
an algorithm such as Bellman-Ford.

@@

\begin{figure}[t]
  \centering
  \begin{minipage}{.49\textwidth}
    \centering
    \begin{tikzpicture}
      \node[draw] (parent) at (0, 0.5) {$\mathsf{parent}$};
      \node[draw] (mother) at (0, -0.5) {$\mathsf{mother}$};
      \node[draw] (sibling) at (2, 0.5) {$\mathsf{sibling}$};
      \node[draw] (father) at (2, -0.5) {$\mathsf{father}$};
      \draw[-{Stealth[scale=1.5]}] (parent) edge node[above] {$+$} (sibling);
      \draw[-{Stealth[scale=1.5]}] (parent) edge node[above] {$+$} (father);
      \draw[-{Stealth[scale=1.5]}] (mother) edge node[below] {$-$} (father);
    \end{tikzpicture}%
    \captionof{figure}{The predicate dependency graph of the program from
      \cref{ex:program}. Positive edges are labelled with `$+$', and negative
      edges with `$-$'.}
    \label{fig:predicate_dependencies}
  \end{minipage}
  \begin{minipage}{.49\textwidth}
    \centering
    \captionof{table}{Types of (potential) dependencies of a predicate
      $\mathsf{P}$ based on the number of undetermined edges on the path from
      the dependency to $\mathsf{P}$}
    \label{tbl:dependencies}
    \begin{tabular}{clc}
      \toprule
      Edges & Name & Notation \\
      \midrule
      $0$ & determined & $\Determined(p)$ \\
      $1$ & almost determined & $\AlmostDetermined(p, s, t)$ \\
      $>1$ & undetermined & $\Undetermined(p)$ \\
      \bottomrule
    \end{tabular}
  \end{minipage}
\end{figure}

@@

  \begin{figure}[t]
    \begin{minipage}{.49\textwidth}
      \centering
      \subfloat[The adjacency matrix of the graph. The boxed value is the
      decision variable that will be propagated by
      \cref{alg:independence_propagation}.\label{fig:dependencies_matrix}]{
        $\begin{blockarray}{lcccc}
          \begin{block}{l!{\quad}(cccc)}
            \mathsf{father} & 0 & 0 & 0 & 0 \\
            \mathsf{mother} & 1 & 0 & 0 & 0 \\
            \mathsf{parent} & 1 & \fbox{\{ 0, 1 \}} & \{ 0, 1 \} & \{ 0, 1 \} \\
            \mathsf{sibling} & 0 & 0 & 0 & 0 \\
          \end{block}
        \end{blockarray}$
      }
    \end{minipage}
    \hfill
    \begin{minipage}{.49\textwidth}
      \centering
      \subfloat[A drawing of the graph. Dashed edges are undetermined---they may
      or may not exist.\label{fig:dependencies2}]{
        \centering
        \makebox[\linewidth]{
          \begin{tikzpicture}
            \node[draw] (parent) at (0, 0.5) {$\mathsf{parent}$};
            \node[draw] (mother) at (0, -0.5) {$\mathsf{mother}$};
            \node[draw] (sibling) at (2, 0.5) {$\mathsf{sibling}$};
            \node[draw] (father) at (2, -0.5) {$\mathsf{father}$};
            \draw[-{Stealth}] (parent) edge (father);
            \draw[-{Stealth}] (mother) edge (father);
            \draw[-{Stealth},dashed,darkgray] (parent) edge (sibling);
            \draw[-{Stealth},dashed,darkgray] (parent) edge (mother);
          \end{tikzpicture}
        }
      }
    \end{minipage}
    \caption{The predicate dependency graph of \cref{example:independence}}
    \label{fig:example}
  \end{figure}

@@

We now present the results of two experiments: in \cref{sec:experiment1} we
examine the scalability of our constraint model with respect to its parameters
and in \cref{sec:experiment2} we demonstrate how the model can be used to
compare inference algorithms and describe their behaviour across a wide range of
programs. The experiments were run on a system with Intel Core i5-8250U
processor, \SI{8}{\giga\byte} RAM, and Arch Linux~5.6.11-arch1-1 operating
system. The constraint model was implemented in Java~8 with
Choco~4.10.2 \cite{choco}. All inference algorithms are implemented in
\textsc{ProbLog}~2.1.0.39 and were run using Python~3.8.2 with PySDD~0.2.10 and
PyEDA~0.28.0. For both sets of experiments, we generate programs without
negative cycles and use a \SI{60}{\second} timeout.

@@

Along with constraints, variables, and their domains, two more design decisions
are needed to complete the model: heuristics and restarts. By trial and error,
the variable ordering heuristic was devised to eliminate sources of
\emph{thrashing}, i.e., situations where a contradiction is being `fixed' by
making changes that have no hope of fixing the contradiction. Thus, we partition
all decision variables into an ordered list of groups and require the values of
all variables from one group to be determined before moving to the next group.
Within each group, we use the `fail first' variable ordering heuristic. The
first group consists of all head predicates. Afterwards, we handle all remaining
decision variables from the first clause before proceeding to the next. The
decision variables within each clause are divided into
\begin{enumerate*}[(a)]
\item the $\variable{structure}$ array,
\item body predicates,
\item head arguments,
\item (if $|\variables{}| > 1$) the $\variable{intros}$ array,
\item body arguments.
\end{enumerate*}
For instance, in the clause from \cref{example:sibling}, all visible parts of
the clause would be decided in this order:
\[
  \overset{1}{\mathsf{sibling}}(\overset{3}{X}, \overset{3}{Y}) \gets
  \overset{2}{\mathsf{parent}}(\overset{4}{X}, \overset{4}{Z})
  \overset{2}{\land} \overset{2}{\mathsf{parent}}(\overset{4}{Y},
  \overset{4}{Z}).
\]
We also employ a geometric restart policy, restarting after $10, 10 \times 1.1,
10 \times 1.1^2, \dots$ contradictions.\footnote{Restarts help overcome early
  mistakes in the search process but can be disabled if one wants to find all
  solutions, in which case search is complete regardless of the variable
  ordering heuristic.} We ran close to \num{400000} experiments, investigating
whether the model is efficient enough to generate reasonably-sized programs and
gaining insight into what parameter values make the CSP harder. For
$|\predicates{}|$, $|\variables{}|$, $|\constants{}|$, $\maxNumNodes{}$, and
$\maxNumClauses{} - |\predicates{}|$ (i.e., the number of clauses in addition to
the mandatory $|\predicates{}|$ clauses), we assign all combinations of 1, 2, 4,
8. $\maxArity{}$ is assigned to values 1--4. For each $|\predicates{}|$, we also
iterate over all possible numbers of independent pairs of predicates, ranging
from 0 up to $\binom{|\predicates{}|}{2}$. For each combination of the
above-mentioned parameters, we pick ten random ways to assign arities to
predicates (such that $\maxArity{}$ occurs at least once) and ten random
combinations of independent pairs.

>>>>

\documentclass[runningheads]{llncs}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage[capitalise]{cleveref}
\usepackage{tikz}
\usepackage{mathrsfs}
\usepackage[nounderscore]{syntax}
\usepackage{blkarray}
\usepackage[binary-units]{siunitx}
\usepackage[inline,shortlabels]{enumitem}
\usepackage{capt-of}
\usepackage[caption=false]{subfig}
\usepackage{booktabs}
\usepackage[misc,geometry]{ifsym}

\usetikzlibrary{arrows.meta}
\captionsetup[subfigure]{width=\linewidth}
\spnewtheorem{constraint}{Constraint}{\bfseries}{\itshape}

\renewcommand\fbox{\fcolorbox{red}{white}}
\newcommand{\hilight}[1]{\setlength{\fboxsep}{1pt}\colorbox{lightgray}{#1}}

@@

@@

@@

\title{Generating Random Logic Programs Using Constraint Programming}
\author{Paulius Dilkas\inst{1} \and Vaishak Belle\inst{1,2}$^{(\textrm{\Letter})}$}
\authorrunning{P. Dilkas and V. Belle}
\institute{University of Edinburgh, Edinburgh, UK
  \email{\{p.dilkas@sms.,vaishak@\}ed.ac.uk} \and Alan Turing Institute, London,
  UK}

@@

  \keywords{Constraint programming \and Probabilistic logic programming \and
    Statistical relational learning}
\end{abstract}

@@

In the world of CSPs, we also have \emph{(constraint) variables}, each with a
\emph{domain}, whose values are restricted using \emph{constraints}. All
constraint variables in the model are integer or set variables, however, if an
integer refers to a logical construct (e.g., a logical variable or a constant),
we will make no distinction between the two. We say that a constraint variable
is \emph{(fully) determined} if its domain (at the time) has exactly one value.
We let $\Box$ denote the absent/disabled value of an optional variable
\cite{DBLP:conf/cpaior/MearsSSTMW14}. We write $\arrayd{a}{b}{c}$ to mean that
$\variable{a}$ is an array of variables of length $b$ such that each element of
$\variable{a}$ has domain $c$. Similarly, we write $\arrayt{a}{b}{c}$ to denote
an array $\variable{a}$ of length $b$ such that each element of $\variable{a}$
has type $\variable{c}$. Finally, we assume that all arrays start with index
zero.

\paragraph{Parameters of the model.} We begin by defining sets and lists of the
primitives used in constructing logic programs: a list of predicates
$\predicates{}$, a list of their corresponding arities $\arities{}$ (so
$|\arities{}|$ = $|\predicates{}|$), a set of variables $\variables{}$, and a
set of constants $\constants{}$. Either $\variables{}$ or $\constants{}$ can be
empty, but we assume that $|\constants{}| + |\variables{}| > 0$. Similarly, the
model supports zero-arity predicates but requires at least one predicate to have
non-zero arity. For notational convenience, we also set $\maxArity{} = \max
\arities{}$. Next, we need a measure of how complex a body of a clause can be.
As we represent each body by a tree (see \cref{sec:bodies}), we set
$\maxNumNodes{} \ge 1$ to be the maximum number of nodes in the tree
representation of any clause. We also set $\maxNumClauses{}$ to be the maximum
number of clauses in a program. We must have that $\maxNumClauses{} \ge
|\predicates{}|$ because we require each predicate to have at least one clause
that defines it. The model supports enforcing predicate independence (see
\cref{sec:independence}), so a set of independent pairs of predicates is another
parameter. Since this model can generate probabilistic as well as
non-probabilistic programs, each clause is paired with a probability which is
randomly selected from a given list---our last parameter. For generating
non-probabilistic programs, one can set this list to $[1]$. Finally, we define
$\tokens{} = \{ \neg, \land, \lor, \top \}$ as the set of tokens that (together
with atoms) form a clause. All decision variables of the model can now be
divided into $2 \times \maxNumClauses{}$ separate groups, treating the body and
the head of each clause separately. We say that the variables are contained in
two arrays: $\arrayt{bodies}{\maxNumClauses{}}{Body}$ and
$\arrayt{heads}{\maxNumClauses{}}{Head}$.

@@

We define the \emph{head} of a clause as a $\variable{predicate} \in \predicates{}
\cup \{ \Box \}$ and $\arrayd{arguments}{\maxArity{}}{\constants{} \cup
  \variables{}} \cup \{ \Box \}$. Here, we use $\Box$ to denote either a
disabled clause that we choose not to use or disabled arguments if the arity of
the $\variable{predicate}$ is less than $\maxArity{}$. The reason why we need a
separate value for the latter (i.e., why it is not enough to fix disabled
arguments to a single already-existing value) will become clear in
\cref{sec:variable_symmetry}. This $\variable{predicate}$ variable has a
corresponding arity that depends on the $\variable{predicate}$. We can
define $\variable{arity} \in [0, \maxArity{}]$ as the arity of the
$\variable{predicate}$ if $\variable{predicate} \in \predicates{}$ and zero
otherwise using the table constraint \cite{DBLP:conf/cpaior/MairyDL15}. This
constraint uses a set of pairs of the form $(p, a)$, where $p$ ranges over all
possible values of the $\variable{predicate}$, and $a$ is either the arity of
predicate $p$ or zero. Having defined arity, we can now fix the superfluous
arguments.
\begin{constraint} \label{constr:arity}
  For $i = 0, \dots, \maxArity{} - 1$, $\variable{arguments}[i] = \Box \iff i
  \ge \variable{arity}$.
\end{constraint}
We also add a constraint that each predicate should get at least one clause.
\begin{constraint}
  Let $P = \{ h.\variable{predicate} \mid h \in \variable{heads} \}$ be a
  multiset. Then
  \[
    \variable{nValues}(P) = \begin{cases}
      |\predicates{}| & \text{if } \variable{count}(\Box, P) = 0 \\
      |\predicates{}| + 1 & \text{otherwise,}
    \end{cases}
  \]
  where $\variable{nValues}(P)$ counts the number of unique values in $P$, and
  $\variable{count}(\Box, P)$ counts how many times $\Box$ appears in $P$.
\end{constraint}
Finally, we want to disable duplicate clauses but with one exception: there may
be more than one disabled clause, i.e., a clause with head $\variable{predicate}
= \Box$. Assuming a lexicographic order over entire clauses such that $\Box >
\mathsf{P}$ for all $\mathsf{P} \in \predicates{}$ and the head predicate is the
`first digit' of this representation, the following constraint disables
duplicates as well as orders the clauses.
\begin{constraint}
  For $i = 1, \dots, \maxNumClauses{} - 1$, if $\variable{heads}[i].\variable{predicate}
  \ne \Box$, then
  \[
    (\variable{heads}[i-1], \variable{bodies}[i-1]) < (\variable{heads}[i],
    \variable{bodies}[i]).
  \]
\end{constraint}

@@

To demonstrate the correctness of the model, this section derives combinatorial
expressions for counting the number of programs with up to $\maxNumClauses{}$
clauses and up to $\maxNumNodes{}$ nodes per clause, and arbitrary
$\predicates{}$, $\arities{}$, $\variables{}$, and $\constants{}$. Being able to
establish two ways to generate the same sequence of numbers (i.e., numbers of
programs with certain properties and parameters) allows us to gain confidence
that the constraint model accurately matches our intentions. For this section,
we introduce the term \emph{total arity} of a body of a clause to refer to the
sum total of arities of all predicates in the body.

@@

Let us order the elements of $\predicates{}$, and let $a_i$ be the arity of the
$i$th predicate. The number of programs is then:
\begin{equation} \label{eq:counting}
  \sum_{\substack{ \sum_{i=1}^{|\predicates{}|} h_i = n,\\
      |\predicates{}| \le n \le \maxNumClauses{},\\
      h_i \ge 1 \text{ for all } i}} \prod_{i=1}^{|\predicates{}|}
  \binom{\sum_{a=0}^{\maxArity{} \times \maxNumNodes{}} C(a) P(a+a_i)}{h_i},
\end{equation}
Here, we sum over all ways to distribute $|\predicates{}| \le n \le
\maxNumClauses{}$ clauses among $|\predicates{}|$ predicates so that each
predicate gets at least one clause. For each predicate, we can then count the
number of ways to select its clauses out of all possible clauses. The number of
possible clauses can be computed by considering each possible arity $a$, and
multiplying the number of `unfinished' clauses $C(a)$ by the number of ways to
select the required $a+a_i$ terms in the body and the head of the clause.
Finally, we compare the numbers produced by \eqref{eq:counting} with the numbers
of programs generated by our model in \num{1032} different scenarios, thus
showing that the combinatorial description developed in this section matches the
model's behaviour.

@@

\emph{Stratification} is a condition necessary for probabilistic logic programs
\cite{DBLP:conf/padl/MantadelisR17} and often enforced on logic programs
\cite{DBLP:journals/tcs/Bidoit91} that helps to ensure a unique answer to every
query. This is achieved by restricting the use of negation so that any program
$\mathscr{P}$ can be partitioned into a sequence of programs $\mathscr{P} =
\bigsqcup_{i=1}^n \mathscr{P}_i$ such that, for all $i$, the negative literals
in $\mathscr{P}_i$ can only refer to predicates defined in $\mathscr{P}_j$ for
$j \le i$ \cite{DBLP:journals/tcs/Bidoit91}.

\emph{Independence}, on the other hand, is defined on a pair of predicates (say,
$\mathsf{P}, \mathsf{Q} \in \mathcal{P}$) and can be interpreted in two ways.
First, if $\mathsf{P}$ and $\mathsf{Q}$ are independent, then any ground atom of
$\mathsf{P}$ is independent of any ground atom of $\mathsf{Q}$ in the underlying
probability distribution of the probabilistic program. Second, the part of the
program needed to fully define $\mathsf{P}$ is disjoint from the part of the
program needed to define $\mathsf{Q}$.

These two seemingly disparate concepts can be defined using the same building
block, i.e., a predicate dependency graph. Let $\mathscr{P}$ be a probabilistic
logic program with its set of predicates $\predicates{}$. Its \emph{(predicate)
  dependency graph} is a directed graph $G_{\mathscr{P}}$ with elements of
$\predicates{}$ as nodes and an edge between $\mathsf{P}, \mathsf{Q} \in
\predicates{}$ if there is a clause in $\mathscr{P}$ with $\mathsf{Q}$ as the
head and $\mathsf{P}$ mentioned in the body. We say that the edge is
\emph{negative} if there exists a clause with $\mathsf{Q}$ as the head and at
least one instance of $\mathsf{P}$ at the body such that the path from the root
to the $\mathsf{P}$ node in the tree representation of the clause passes through
at least one negation node; otherwise, it is \emph{positive}. We say that
$\mathscr{P}$ (or $G_{\mathscr{P}}$) has a \emph{negative cycle} if
$G_{\mathscr{P}}$ has a cycle with at least one negative edge. A program
$\mathscr{P}$ is \emph{stratified} if $G_{\mathscr{P}}$ has no negative
cycles.\footnote{This definition is an extension of a well-known result for
  logic programs \cite{DBLP:journals/jlp/BalbinPRM91} to probabilistic logic
  programs with arbitrary complex clause bodies.} Thus a simple entailment
algorithm for stratification can be constructed by selecting all clauses, all
predicates of which are fully determined, and looking for negative cycles in the
dependency graph constructed based on those clauses using an algorithm such as
Bellman-Ford.

@@

\begin{figure}[t]
  \centering
  \begin{minipage}{.49\textwidth}
    \centering
    \begin{tikzpicture}
      \node[draw] (parent) at (0, 0.5) {$\mathsf{parent}$};
      \node[draw] (mother) at (0, -0.5) {$\mathsf{mother}$};
      \node[draw] (sibling) at (2, 0.5) {$\mathsf{sibling}$};
      \node[draw] (father) at (2, -0.5) {$\mathsf{father}$};
      \draw[-{Stealth[scale=1.5]}] (parent) edge node[above] {$+$} (sibling);
      \draw[-{Stealth[scale=1.5]}] (parent) edge node[above] {$+$} (father);
      \draw[-{Stealth[scale=1.5]}] (mother) edge node[below] {$-$} (father);
    \end{tikzpicture}%
    \captionof{figure}{The predicate dependency graph of the program from
      \protect{\cref{ex:program}}. Positive edges are labelled with `$+$', and
      negative edges with `$-$'.}
    \label{fig:predicate_dependencies}
  \end{minipage}
  \begin{minipage}{.49\textwidth}
    \centering
    \captionof{table}{Types of (potential) dependencies of a predicate
      $\mathsf{P}$ based on the number of undetermined edges on the path from
      the dependency to $\mathsf{P}$}
    \label{tbl:dependencies}
    \begin{tabular}{clc}
      \toprule
      Edges & Name & Notation \\
      \midrule
      $0$ & determined & $\Determined(p)$ \\
      $1$ & almost determined & $\AlmostDetermined(p, s, t)$ \\
      $>1$ & undetermined & $\Undetermined(p)$ \\
      \bottomrule
    \end{tabular}
  \end{minipage}
\end{figure}

@@

  \begin{figure}[t]
    \begin{minipage}{.49\textwidth}
      \centering
      \subfloat[The adjacency matrix of the graph. The boxed value is the
      decision variable that will be propagated by
      \cref{alg:independence_propagation}.\label{fig:dependencies_matrix}]{
        $\begin{blockarray}{lcccc}
          \begin{block}{l!{\quad}(cccc)}
            \mathsf{father} & 0 & 0 & 0 & 0 \\
            \mathsf{mother} & 1 & 0 & 0 & 0 \\
            \mathsf{parent} & 1 & \fbox{\{ 0, 1 \}} & \{ 0, 1 \} & \{ 0, 1 \} \\
            \mathsf{sibling} & 0 & 0 & 0 & 0 \\
          \end{block}
        \end{blockarray}$
      }
    \end{minipage}
    \hfill
    \begin{minipage}{.49\textwidth}
      \centering
      \subfloat[A drawing of the graph. Dashed edges are undetermined---they may
      or may not exist.\label{fig:dependencies2}]{
        \centering
        \makebox[\linewidth]{
          \begin{tikzpicture}
            \node[draw] (parent) at (0, 0.5) {$\mathsf{parent}$};
            \node[draw] (mother) at (0, -0.5) {$\mathsf{mother}$};
            \node[draw] (sibling) at (2, 0.5) {$\mathsf{sibling}$};
            \node[draw] (father) at (2, -0.5) {$\mathsf{father}$};
            \draw[-{Stealth}] (parent) edge (father);
            \draw[-{Stealth}] (mother) edge (father);
            \draw[-{Stealth},dashed,darkgray] (parent) edge (sibling);
            \draw[-{Stealth},dashed,darkgray] (parent) edge (mother);
          \end{tikzpicture}
        }
      }
    \end{minipage}
    \caption{The predicate dependency graph of \protect{\cref{example:independence}}}
    \label{fig:example}
  \end{figure}

@@

We now present the results of two experiments: in \cref{sec:experiment1} we
examine the scalability of our constraint model with respect to its parameters
and in \cref{sec:experiment2} we demonstrate how the model can be used to
compare inference algorithms and describe their behaviour across a wide range of
programs. The experiments were run on a system with Intel Core i5-8250U
processor and \SI{8}{\giga\byte} of RAM. The constraint model was implemented in
Java~8 with Choco~4.10.2 \cite{choco}. All inference algorithms are implemented
in \textsc{ProbLog}~2.1.0.39 and were run using Python~3.8.2 with PySDD~0.2.10
and PyEDA~0.28.0. For both sets of experiments, we generate programs without
negative cycles and use a \SI{60}{\second} timeout.

@@

Along with constraints, variables, and their domains, two more design decisions
are needed to complete the model: heuristics and restarts. By trial and error,
the variable ordering heuristic was devised to eliminate sources of
\emph{thrashing}, i.e., situations where a contradiction is being `fixed' by
making changes that have no hope of fixing the contradiction. Thus, we partition
all decision variables into an ordered list of groups and require the values of
all variables from one group to be determined before moving to the next group.
Within each group, we use the `fail first' variable ordering heuristic. The
first group consists of all head predicates. Afterwards, we handle all remaining
decision variables from the first clause before proceeding to the next. The
decision variables within each clause are divided into
\begin{enumerate*}[(a)]
\item the $\variable{structure}$ array,
\item body predicates,
\item head arguments,
\item (if $|\variables{}| > 1$) the $\variable{intros}$ array,
\item body arguments.
\end{enumerate*}
For instance, in the clause from \cref{example:sibling}, all visible parts of
the clause would be decided in this order:
\[
  \overset{1}{\mathsf{sibling}}(\overset{3}{X}, \overset{3}{Y}) \gets
  \overset{2}{\mathsf{parent}}(\overset{4}{X}, \overset{4}{Z})
  \overset{2}{\land} \overset{2}{\mathsf{parent}}(\overset{4}{Y},
  \overset{4}{Z}).
\]
We also employ a geometric restart policy, restarting after $10, 10 \times 1.1,
10 \times 1.1^2, \dots$ contradictions.\footnote{Restarts help overcome early
  mistakes in the search process but can be disabled if one wants to find all
  solutions, in which case search is complete regardless of the variable
  ordering heuristic.} We ran \num{399360} experiments, investigating the
model's efficiency and gaining insight into what parameter values make the CSP
harder. For $|\predicates{}|$, $|\variables{}|$, $|\constants{}|$,
$\maxNumNodes{}$, and $\maxNumClauses{} - |\predicates{}|$ (i.e., the number of
clauses in addition to the mandatory $|\predicates{}|$ clauses), we assign all
combinations of 1, 2, 4, 8. $\maxArity{}$ is assigned to values 1--4. For each
$|\predicates{}|$, we also iterate over all possible numbers of independent
pairs of predicates, ranging from 0 up to $\binom{|\predicates{}|}{2}$. For each
combination of the above-mentioned parameters, we pick ten random ways to assign
arities to predicates (such that $\maxArity{}$ occurs at least once) and ten
random combinations of independent pairs.
